---
title: 'Effect of Missing Not at Random on Causal Inference'
date: today
date-format: long
format: 
  baylor_theme-revealjs:
    author: ''
    footer: 'Nathen Byford'
---

```{r}
#| include: false

library(tidyverse); theme_set(theme_bw())
library(gtsummary)

longi_score <- read_csv("data/longi_score_real_data.csv")

complete_case <- read_csv("data/complete_case.csv")
```


## Types of Missingness

3 Types of Missingness commonly observed:

1. Missing Completely at Random (MCAR)
2. Missing at Random (MAR)
3. Missing Not at Random (MNAR)

- Best case is MCAR

# Motivating Example

## Clinical Trial with dropout

- Clinical trial data for depression medication
    - Missing data not at random
- Do Causal Inference methods provide biased results for complete case analysis compared to imputation?

# Exploritory Data Analysis

## Response overtime

```{r}
longi_score |> 
  mutate(
    "Incomplete" = ifelse(id %in% complete_case$id, FALSE, TRUE)
  ) |> 
  ggplot(aes(x = week, y = score)) +
  geom_jitter(aes(shape = Incomplete, color = Incomplete), width = .15) +
  geom_smooth() +
  facet_wrap(~treat) +
  scale_color_manual(values = c("black", "red")) +
  scale_shape_manual(values = c(16, 4)) +
  labs(title = "Weekly Response with all Observed Values") +
  theme(legend.position = "bottom")
```

## Treatment

```{r}
longi_score |> 
  pivot_longer(
    cols = c(base, `HAMA TOTAL`, cfb),
    names_to = "measure",
    values_to = "value") |> 
  ggplot(aes(x = value, group = treat)) +
  geom_density(aes(fill = treat), alpha = .9) +
  facet_wrap(~measure)
```


# Complete case Analysis

## Difference in Means


## Difference-in-Difference (DID)

Using DID we can estimate the causal effect using the differences of treated and untreated at week 1 and week 6.

- The DID estimator is not significant with an estimated 95% CI of -1.0 to 0.36

```{r}
#| tbl-cap: "Difference in difference regression estimates"

complete_data <- complete_case |> 
  filter(week %in% c(1, 6))

did_reg <- lm(score ~ treat + week + DID, 
              data = complete_data)

tbl_regression(did_reg)
```


## IP-weighting

```{r}
#| eval: false
library(gfoRmula)

data_cont_time <- complete_case |> 
  mutate(week = case_when(
    week == 4 ~ 3,
    week == 6 ~ 4,
    .default = week
  ))

gfoRmula::gformula(
  obs_data = data_cont_time,
  id = "id",
  time_name = "week",
  covnames = c("hamatotl_locf", "cfb"),
  covparams = list(
    hamatotl_locf ~ base + week,
    cfb ~ base + week
  ),
  covtypes = c("normal", "normal"),
  outcome_name = "score",
  outcome_type = "continuous_eof",
  ymodel = score ~ treat + week + hamatotl_locf + cfb,
  seed = 1234
)

```


```{r}

ps_mod <- complete_case |> 
  filter(week %in% c(1, 6)) |> 
  glm(treat ~ sex + cfb + base, family = "binomial", data = _)

ps <- as.numeric(predict(ps_mod, type = "response"))

wls <- lm(score ~ treat + week + DID, 
              data = complete_data, weights = ps)

tbl_regression(wls)
```

# Imputed Data


## Imputed values

```{r}
imput_data <- read_csv("data/pred_real_data.csv")

imputed_values <- imput_data |> 
  summarise(
    value = mean(predict),
    .by = c(id, week)
  )

new_data <- left_join(longi_score, imputed_values, by = c("id", "week")) |> 
  mutate(score = if_else(is.na(score), value, score)) |> 
  select(-value)

new_data |> 
  ggplot(aes(x = week, y = score)) +
  geom_jitter(width = .15) +
  geom_smooth(aes(color = "Imputed"), se = FALSE) +
  geom_smooth(data = longi_score, se = FALSE, aes(color = "Complete case")) +
  facet_wrap(~treat) +
  labs(title = "Weekly Response with Imputed Values",
       color = "Line") +
  scale_color_manual(values = c("cornflowerblue", "forestgreen")) +
  theme(legend.position = "bottom")
```

## Imputed DID

Using DID we can estimate the causal effect using the differences of treated and untreated at week 1 and week 6.

- The DID estimator is not significant with an estimated 95% CI of -1.0 to 0.36

```{r}
#| tbl-cap: "Difference in difference regression estimates with imputed values"



did_reg <- new_data |> 
  filter(week %in% c(1, 6)) |>
  mutate("DID" = treat * week) |> 
  lm(score ~ treat + week + DID, 
              data = _)

tbl_regression(did_reg)
```

## Imputed IP-weighting

```{r}

ps_mod_i <- new_data |> 
  filter(week %in% c(1, 6)) |> 
  glm(treat ~ sex + base, family = "binomial", data = _)

ps_i <- as.numeric(predict(ps_mod_i, type = "response"))

wls_i <- new_data |> 
  filter(week %in% c(1, 6)) |> 
  mutate("DID" = treat * week) |> 
  lm(score ~ treat + week + DID, 
              data = _, weights = ps_i)

tbl_regression(wls_i)
```


# Observational Study

## Framingham Heart Study

- Long term prospective study of cardiovascular disease. 
- 5,209 subject initially enrolled in the study

```{r}
library(riskCommunicator)
data("framingham")


framingham |> 
  summarise(
    count = n(),
    .by = PERIOD
  ) |> 
  arrange(PERIOD) |> 
  rename(Exam = PERIOD) |> 
  gt::gt()
```

## DID

- Q: Does Smoking effect the probability of Cardiovascular Disease?

```{r}
fram_did <- framingham |> 
  filter(PERIOD %in% c(1, 2)) |> 
  mutate(DID = PERIOD * CURSMOKE) |> 
  glm(CVD ~ CURSMOKE + PERIOD + DID, data = _, family = "binomial")

tbl_regression(fram_did, exponentiate = TRUE)
```

## Imputed Framingham

```{r}
fram_12 <- framingham |> 
  filter(PERIOD %in% c(1, 2))

all_patient_periods <- expand.grid(unique(fram_12$RANDID), unique(fram_12$PERIOD))

# nrow(all_patient_periods) - nrow(fram_12)


```


# Simulated Data

## Simulated data Missingness

:::{.columns}

::::{.column width="70%"}

- Simulated data for a 16 week study with bi-weekly checkups
    - No missing values until week 10
    - Each week of missing values was split 45:75 between treated and untreated

```{r}
#| eval: false


longi_sim |> 
  summarise(
    `NA count` = sum(is.na(score)),
    .by = treat
  ) |> 
  gt::gt()
```

::::

::::{.column width="30%"}
```{r}
longi_sim <- read_csv("data/longi_score_sim.csv")

longi_sim |> 
  summarise(
    `NA count` = sum(is.na(score)),
    .by = week
  ) |> 
  rename("Week" = week) |> 
  gt::gt()
```
::::

:::

## Plots

```{r}
longi_sim |> 
  ggplot(aes(x = week, y = score)) +
  geom_jitter(width = .25) + 
  geom_smooth(method = "loess") +
  facet_wrap(~treat)
```


## Difference-in-Difference (DID)

Using DID we can estimate the causal effect using the differences of treated and untreated at week 1 and week 6.

- The DID estimator is not significant with an estimated 95% CI of -1.0 to 0.36

```{r}

complete_case_sim <- read_csv("data/complete_sim.csv")

complete_sim <- complete_case_sim |> 
  filter(week %in% c(0, 16))

did_sim <- lm(score ~ treat + week + DID, 
              data = complete_sim)

tbl_regression(did_sim)
```


## IP-weighting

```{r}

ps_mod_sim <- complete_sim |> 
  filter(week %in% c(0, 16)) |> 
  glm(treat ~ sex + weight + base, family = "binomial", data = _)

ps_sim <- as.numeric(predict(ps_mod_sim, type = "response"))

wls_sim <- complete_sim |> 
  filter(week %in% c(0, 16)) |> 
  lm(score ~ treat + week + DID, 
              data = _, weights = ps_sim)

tbl_regression(wls_sim)
```

## Imputed values

```{r}
imput_sim <- read_csv("data/pred_sim.csv")

imputed_values <- imput_data |> 
  summarise(
    value = mean(predict),
    .by = c(id, week)
  )

new_data <- left_join(longi_score, imputed_values, by = c("id", "week")) |> 
  mutate(score = if_else(is.na(score), value, score)) |> 
  select(-value)

new_data |> 
  ggplot(aes(x = week, y = score)) +
  geom_jitter(width = .15) +
  geom_smooth(aes(color = "Imputed"), se = FALSE) +
  geom_smooth(data = longi_score, se = FALSE, aes(color = "Complete case")) +
  facet_wrap(~treat) +
  labs(title = "Weekly Response with Imputed Values",
       color = "Line") +
  scale_color_manual(values = c("cornflowerblue", "forestgreen")) +
  theme(legend.position = "bottom")
```

## Imputed DID

Using DID we can estimate the causal effect using the differences of treated and untreated at week 1 and week 6.

- The DID estimator is not significant with an estimated 95% CI of -1.0 to 0.36

```{r}
#| tbl-cap: "Difference in difference regression estimates with imputed values"



did_reg <- new_data |> 
  filter(week %in% c(1, 6)) |>
  mutate("DID" = treat * week) |> 
  lm(score ~ treat + week + DID, 
              data = _)

tbl_regression(did_reg)
```

## Imputed IP-weighting

```{r}

ps_mod_i <- new_data |> 
  filter(week %in% c(1, 6)) |> 
  glm(treat ~ sex + base, family = "binomial", data = _)

ps_i <- as.numeric(predict(ps_mod_i, type = "response"))

wls_i <- new_data |> 
  filter(week %in% c(1, 6)) |> 
  mutate("DID" = treat * week) |> 
  lm(score ~ treat + week + DID, 
              data = _, weights = ps_i)

tbl_regression(wls_i)
```
